\bf Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-06-13 pkg/paper/Paper/JAE.Rnw
updated JAE
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@123 b679b2f7-6539-45d0-a605-5ddcb976cde9

<<slovenia-plots2, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
for(i in 1:2) plot(x$model[[i]], moments = TRUE,
  xlim = c(-1.5, 2.2), ylim = c(0, 1.3),
  breaks = seq(-1.5, 2, by = 0.25),
  xlab = paste(names(x$model)[i], "\n", "Goodness-of-fit p-value: ",
    format.pval(summary(x$model[[i]])$chisq.test$p.value, digits = 4),
    sep = ""))
@
\caption{\label{fig:slovenia2} Histogram of observed monthly inflation rates in
Slovenia for \Sexpr{names(x$model)[1]} (left) and \Sexpr{names(x$model)[1]} (right),
along with fitted generalized logistic probability density function, associated
moments, and $\chi^2$ goodness-of-fit test.}
\end{figure}

@@

As there is evidence for at least one break, the LWZ criterion is employed for $B = 1, 2, \dots$
breakpoints with a minimal segment size of two years per segment. It assumes its minimum for $B = 1$
associated with the breakpoint \Sexpr{breakdates(x$breakpoints, breaks = x$nbreaks)}, depicted
by a vertical dashed line in both panels of Figure~\ref{fig:slovenia1}. The resulting segmented
fitted mean (horizontal blue lines in the left panel Figure~\ref{fig:slovenia1}) shows that
before the break, Slovenia experienced very high inflation rates but was successful in reducing
inflation to a much lower level afterwards.

@@

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
late 1990s or early 2000s. Czech Republic, Estonia and Hungary experienced a break in 1997/1998
resulting from efforts to curb in inflation by decreasing the growth of the money 
supply. The timing of the break thus indicates the date when the the biggest
transitionally shocks on the way towards free market economies have been overcome. In Poland, this
transition took two more years. Slovakia and Slovenia are somewhat different: On the one hand,
they started with lower mean inflation rates and on the other hand, their inflation dynamics
both changed roughly one year prior to their entry into the ERM~II.

@@

The other countries do not exhibit any significant breakpoints during the
mid-1990s, but many changes occur in the transition period from \Sexpr{yrs[2]}
to \Sexpr{yrs[3]}, encompassing both the completion of transformation from
planned to free market economies and the intended entry into the EMU.
During this time, the countries were successful in
decreasing all three moments of their inflation distributions while
simultaneously decreasing the heterogeneity (as captured by the corresponding
interquartile ranges) across countries. Before the transition, all three
moments were extremely large compared to the other two groups, while afterwards
all moments are much closer to those of the other groups. In Jan~\Sexpr{yrs[3]},
mean inflation is in between the levels for the core euro and PIIGS countries,
variance is still slightly higher, and skewness is not too far away from symmetry.

@@

More specifically, the mean inflation of the euro area  from a GL fit in the 
year before the crisis (\Sexpr{yearmon(2007 + 7/12)}--\Sexpr{yearmon(2008 + 6/12)}) was
$\Sexpr{round3(glmean(window(eu, start = yearmon(2007 + 7/12), end = yearmon(2008 + 6/12))))}$
and after the crisis (\Sexpr{yearmon(2008 + 7/12)}--\Sexpr{yearmon(2009 + 6/12)}) it dropped to
$\Sexpr{round3(glmean(window(eu, start = yearmon(2008 + 7/12), end = yearmon(2009 + 6/12))))}$
before rising again in the following year  (\Sexpr{yearmon(2009 + 7/12)}--\Sexpr{yearmon(2010 + 6/12)}):
$\Sexpr{round3(glmean(window(eu, start = yearmon(2009 + 7/12), end = yearmon(2010 + 6/12))))}$.
One way to bring out this pattern in graphically is
Figure~\ref{fig:glmean}. It depicts the inflation rates (in gray) and a 12-month rolling mean (in red),
calculated again as the first moment of a fitted GL distribution.
The strong increase in mean inflation during the boom years of 2007/2008
accompanied by an equally strong deflation following the burst of the bubble
is clearly visible for most countries. However, the level of mean inflation
did not change persistently, as the inflation rates seem to have returned to
their prior path, conveying the same message as the inflation series of the
euro area.

@@

Our results were obtained using 
\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \citep{R} with the packages
\pkg{glogis}~\Sexpr{packageDescription("glogis")["Version"]},
\pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \citep{Zeileis+Leisch+Hornik:2002}, and
\pkg{fxregime}~\Sexpr{packageDescription("fxregime")["Version"]} \citep{Zeileis+Shah+Patnaik:2010},
all of which are available under the General Public License (GPL) from the
Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}). Package \pkg{strucchange} provides
various techniques for inference in structural-change settings which are complemented
with additional maximum-likelihood-based methods by \pkg{fxregime}. Inference for the
generalized logistic model (that can be combined with \pkg{strucchange}/\pkg{fxregime})
is implemented in the new package \pkg{glogis} that also contains the data used, both
the raw monthly price indexes \citep[as provided by][]{oecd} and the seasonally adjusted
monthly inflation rates. The seasonal adjustment was performed in \pkg{X-12-ARIMA}~0.3
(\citealp{X-12-ARIMA}, see also \citealp{Findley+Monsell+Bell:1998}) through the interface
provided by \proglang{gretl}~1.9.5 (\citealp{gretl}, see also \citealp{Smith+Mixon:2006} for a review).

>>>>

<<slovenia-plots2, fig=TRUE, height=4.2, width=9.6>>=
par(mfrow = c(1, 2))
for(i in 1:2) plot(x$model[[i]], moments = TRUE,
  xlim = c(-1.5, 2.2), ylim = c(0, 1.3),
  breaks = seq(-1.5, 2, by = 0.25),
  xlab = paste(names(x$model)[i], "\n", "Goodness-of-fit p-value: ",
    format.pval(summary(x$model[[i]])$chisq.test$p.value, digits = 4),
    sep = ""))
@
\caption{\label{fig:slovenia2} Histogram of observed monthly inflation rates in
Slovenia for \Sexpr{names(x$model)[1]} (left) and \Sexpr{names(x$model)[2]} (right),
along with fitted generalized logistic probability density function, associated
moments, and $\chi^2$ goodness-of-fit test.}
\end{figure}

@@

As there is evidence for at least one break, the LWZ criterion is employed for $B = 1, 2, \dots$
breakpoints with a minimal segment size of two years per segment. It assumes its minimum for $B = 1$
associated with the breakpoint \Sexpr{breakdates(x$breakpoints, breaks = x$nbreaks)}, depicted
by a vertical dashed line in both panels of Figure~\ref{fig:slovenia1}. The resulting segmented
fitted mean (horizontal blue lines in the left panel of Figure~\ref{fig:slovenia1}) shows that
before the break, Slovenia experienced very high inflation rates but was successful in reducing
inflation to a much lower level afterwards.

@@

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
late 1990s or early 2000s. Czech Republic, Estonia, and Hungary experienced a break in 1997/1998
resulting from efforts to curb in inflation by decreasing the growth of the money 
supply. The timing of the break thus indicates the date when the the biggest
transitionally shocks on the way towards free market economies have been overcome. In Poland, this
transition took two more years. Slovakia and Slovenia are somewhat different: On the one hand,
they started with lower mean inflation rates and on the other hand, their inflation dynamics
both changed roughly one year prior to their entry into the ERM~II.

@@

The other countries do not exhibit any significant breakpoints during the
mid-1990s, but many changes occur in the transition period from \Sexpr{yrs[2]}
to \Sexpr{yrs[3]}, encompassing both the completion of the transformation from
planned to free market economies and the intended entry into the EMU.
During this time, the countries were successful in
decreasing all three moments of their inflation distributions while
simultaneously decreasing the heterogeneity (as captured by the corresponding
interquartile ranges) across countries. Before the transition, all three
moments were extremely large compared to the other two groups, while afterwards
all moments are much closer to those of the other groups. In Jan~\Sexpr{yrs[3]},
mean inflation is in between the levels for the core euro and PIIGS countries,
variance is still slightly higher, and skewness is not too far away from symmetry.

@@

More specifically, the mean inflation of the euro area  from a GL fit in the 
year before the crisis (\Sexpr{yearmon(2007 + 7/12)}--\Sexpr{yearmon(2008 + 6/12)}) was
$\Sexpr{round3(glmean(window(eu, start = yearmon(2007 + 7/12), end = yearmon(2008 + 6/12))))}$
and after the crisis (\Sexpr{yearmon(2008 + 7/12)}--\Sexpr{yearmon(2009 + 6/12)}) it dropped to
$\Sexpr{round3(glmean(window(eu, start = yearmon(2008 + 7/12), end = yearmon(2009 + 6/12))))}$
before rising again in the following year  (\Sexpr{yearmon(2009 + 7/12)}--\Sexpr{yearmon(2010 + 6/12)}):
$\Sexpr{round3(glmean(window(eu, start = yearmon(2009 + 7/12), end = yearmon(2010 + 6/12))))}$.
One way to bring out this pattern graphically is
Figure~\ref{fig:glmean}. It depicts the inflation rates (in gray) and a 12-month rolling mean (in red),
calculated again as the first moment of a fitted GL distribution.
The strong increase in mean inflation during the boom years of 2007/2008
accompanied by an equally strong deflation following the burst of the bubble
is clearly visible for most countries. However, the level of mean inflation
did not change persistently, as the inflation rates seem to have returned to
their prior path, conveying the same message as the inflation series of the
euro area.

@@

Our results were obtained using 
\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} \citep{R} with the packages
\pkg{glogis}~\Sexpr{packageDescription("glogis")["Version"]},
\pkg{strucchange}~\Sexpr{packageDescription("strucchange")["Version"]} \citep{Zeileis+Leisch+Hornik:2002}, and
\pkg{fxregime}~\Sexpr{packageDescription("fxregime")["Version"]} \citep{Zeileis+Shah+Patnaik:2010},
all of which are available under the General Public License (GPL) from the
Comprehensive \proglang{R} Archive Network (\url{http://CRAN.R-project.org/}). Package \pkg{strucchange} provides
various techniques for inference in structural-change settings which are complemented
with additional maximum-likelihood-based methods by \pkg{fxregime}. Inference for the
generalized logistic model (that can be combined with \pkg{strucchange}/\pkg{fxregime})
is implemented in the new package \pkg{glogis} that also contains the data used, both
the raw monthly price indexes \citep[as provided by the][]{oecd} and the seasonally adjusted
monthly inflation rates. The seasonal adjustment was performed in \pkg{X-12-ARIMA}~0.3
(\citealp{X-12-ARIMA}, see also \citealp{Findley+Monsell+Bell:1998}) through the interface
provided by \proglang{gretl}~1.9.5 (\citealp{gretl}, see also \citealp{Smith+Mixon:2006} for a review).
